---
title: "Scaling Bundestag Roll-Call Votes: SVD and IRT"
subtitle: "DS 201 Machine Learning and Causal Inference — AI-Forward Track"
author:
  - name: Maximilian Birkle
    affiliation: Mannheim Master in Social Data Science
    student-id: "1831999"
date: today
date-format: "MMMM D, YYYY"

format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: true
    code-fold: true
    code-tools: true
    code-summary: "Show code"
    fig-align: center
    fig-width: 9
    fig-height: 6
    embed-resources: true
    smooth-scroll: true
    highlight-style: github

execute:
  echo: true
  warning: false
  message: false
  cache: true

knitr:
  opts_chunk:
    cache.path: "_cache/"
---

```{r setup}
#| include: false
library(tidyverse)
library(brms)
library(bayesplot)
library(ggrepel)   # non-overlapping labels on plots
library(patchwork) # combining plots side-by-side

# consistent ggplot theme matching bayesplot defaults
theme_set(bayesplot::theme_default(base_size = 12))

# Stan options: compile once, use all cores
rstan::rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())

# cache fitted Stan models so they survive re-renders
if (!dir.exists("models")) dir.create("models")
```

---

> **Track:** AI-Forward — Claude used throughout for code generation and
> explanation.
> **Data:** 20th Bundestag roll-call votes, Oct 2021–Mar 2025,
> sourced from Abgeordnetenwatch e.V. (CC0).

---

# Data

## Load and inspect

```{r load-data}
# ── Replace the filename below once you receive the CSV ──────────────────────
df_raw <- read_csv("data/votes.csv")   # adjust path / filename as needed

glimpse(df_raw)
head(df_raw, 10)
```

## Recode votes

The lecture specifies: Ja → 1, Nein → 0, Enthaltung / Nicht abgegeben → `NA`.
Abstentions are treated as missing rather than recoded to 0 because they are
strategic choices that would bias ideal-point estimates toward the centre.

```{r recode}
# ── Adjust column names to match the actual CSV ──────────────────────────────
# Expected columns (rename as needed):
#   legislator_id  : unique ID per MP
#   legislator_name: display name
#   party          : Fraktion / party group
#   bill_id        : unique ID per roll-call vote
#   vote_raw       : raw vote string ("ja", "nein", "Enthaltung", ...)

df <- df_raw |>
  rename(
    legislator_id   = TODO_legislator_id_col,   # <── fill in
    legislator_name = TODO_name_col,
    party           = TODO_party_col,
    bill_id         = TODO_bill_id_col,
    vote_raw        = TODO_vote_col
  ) |>
  mutate(
    vote = case_when(
      str_to_lower(vote_raw) %in% c("ja", "yes", "1")   ~ 1L,
      str_to_lower(vote_raw) %in% c("nein", "no", "0")  ~ 0L,
      TRUE                                               ~ NA_integer_
    )
  )

# Sanity check: distribution of vote values
df |> count(vote_raw, vote)
```

## Build the vote matrix

```{r build-matrix}
# Wide format: rows = legislators, columns = bills, cells = 0/1/NA
vote_wide <- df |>
  select(legislator_id, bill_id, vote) |>
  pivot_wider(names_from = bill_id, values_from = vote,
              values_fn = first)   # keep first if duplicates exist

# Store legislator metadata for later labelling
leg_meta <- df |>
  distinct(legislator_id, legislator_name, party)

# Numeric matrix (rownames = legislator IDs)
X_raw <- vote_wide |>
  select(-legislator_id) |>
  as.matrix()
rownames(X_raw) <- vote_wide$legislator_id

cat("Matrix dimensions:", nrow(X_raw), "legislators ×", ncol(X_raw), "bills\n")
cat("Missing rate:", round(mean(is.na(X_raw)) * 100, 1), "%\n")
```

---

# Task 1: SVD of the Raw Vote Matrix

## Impute missing values (column-mean rule)

SVD requires a complete matrix.  We impute each missing cell with that bill's
observed yes-rate — equivalent to saying "this MP voted like an average member
of parliament on this bill."

```{r impute-raw}
X_imputed <- X_raw
col_means  <- colMeans(X_raw, na.rm = TRUE)

for (j in seq_len(ncol(X_imputed))) {
  missing_rows <- is.na(X_imputed[, j])
  X_imputed[missing_rows, j] <- col_means[j]
}

cat("Missing cells after imputation:", sum(is.na(X_imputed)), "\n")
```

## Compute SVD

```{r svd-raw}
svd_raw <- svd(X_imputed)

# First singular value's share of total variance
sv_share <- svd_raw$d^2 / sum(svd_raw$d^2)
cat("Variance explained by dim 1:", round(sv_share[1] * 100, 1), "%\n")
cat("Variance explained by dim 2:", round(sv_share[2] * 100, 1), "%\n")
```

## Plot: legislator positions on dimension 1

```{r plot-svd-raw}
#| fig-cap: "Raw SVD — first left-singular vector (legislator ideal points). Higher values = one pole, lower = the other."

svd_raw_scores <- tibble(
  legislator_id = rownames(X_raw),
  dim1          = svd_raw$u[, 1],
  dim2          = svd_raw$u[, 2]
) |>
  left_join(leg_meta, by = "legislator_id")

# Party colour palette — adjust to actual Fraktion names in the data
party_colours <- c(
  "SPD"     = "#E3000F",
  "CDU/CSU" = "#000000",
  "Grüne"   = "#1AA037",
  "FDP"     = "#FFED00",
  "AfD"     = "#009EE0",
  "Linke"   = "#BE3075",
  "BSW"     = "#6A0DAD"
)

p_raw <- ggplot(svd_raw_scores, aes(x = dim1, y = 0, colour = party)) +
  geom_jitter(height = 0.3, alpha = 0.6, size = 1.8) +
  geom_text_repel(
    data = svd_raw_scores |> slice_min(dim1, n = 5) |>
                             bind_rows(slice_max(svd_raw_scores, dim1, n = 5)),
    aes(label = legislator_name),
    size = 3, max.overlaps = 20
  ) +
  scale_colour_manual(values = party_colours, na.value = "grey60") +
  labs(
    title   = "Raw SVD — Dimension 1",
    x       = "First left-singular vector (U₁)",
    y       = NULL,
    colour  = "Party"
  ) +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())

p_raw
```

**Face-validity check:** [*Describe who sits at each extreme once the data are
loaded — e.g., "Die Linke / BSW on the left, AfD on the right" or vice versa.
Does dimension 1 look like a left–right axis, a coalition vs. opposition
divide, or something else?*]

---

# Task 2: SVD of the Double-Centred Matrix

## Fit the two-way additive baseline

The double-centring approach removes row (legislator) and column (bill)
main effects so that the SVD captures only the *interaction* — how much each
MP deviates from their own baseline and from the bill's baseline.

Model: $X_{ij} = \mu + r_i + c_j + \varepsilon_{ij}$ on **observed cells only**.
Missing residuals are set to zero before SVD.

```{r double-centre}
# Convert to long format, keep only observed votes
df_obs <- df |>
  filter(!is.na(vote)) |>
  mutate(
    legislator_id = as.factor(legislator_id),
    bill_id       = as.factor(bill_id)
  )

# Fit additive two-way model (no interaction)
fit_additive <- lm(vote ~ legislator_id + bill_id, data = df_obs)

# Residuals on the observed grid
df_obs <- df_obs |>
  mutate(resid = residuals(fit_additive))

# Pivot residuals to wide matrix; fill missing cells with 0
resid_wide <- df_obs |>
  select(legislator_id, bill_id, resid) |>
  pivot_wider(names_from = bill_id, values_from = resid,
              values_fill = 0)

X_dc <- resid_wide |>
  select(-legislator_id) |>
  as.matrix()
rownames(X_dc) <- as.character(resid_wide$legislator_id)

# Align column order to X_raw (add any missing bills as 0)
missing_bills <- setdiff(colnames(X_raw), colnames(X_dc))
if (length(missing_bills) > 0) {
  X_dc <- cbind(X_dc, matrix(0, nrow = nrow(X_dc),
                              ncol = length(missing_bills),
                              dimnames = list(NULL, missing_bills)))
}
X_dc <- X_dc[, colnames(X_raw)]  # same column order as X_raw
```

## Compute SVD on residual matrix

```{r svd-dc}
svd_dc <- svd(X_dc)

sv_share_dc <- svd_dc$d^2 / sum(svd_dc$d^2)
cat("Variance explained by dim 1 (double-centred):",
    round(sv_share_dc[1] * 100, 1), "%\n")
```

## Plot: comparison of both SVDs

```{r plot-svd-dc}
#| fig-cap: "Comparison of raw SVD (left) vs. double-centred SVD (right). Double-centring removes party-bloc main effects so the first dimension captures voting *deviations*."

svd_dc_scores <- tibble(
  legislator_id = rownames(X_raw),
  dim1          = svd_dc$u[, 1],
  dim2          = svd_dc$u[, 2]
) |>
  left_join(leg_meta, by = "legislator_id")

make_strip_plot <- function(scores_df, title) {
  ggplot(scores_df, aes(x = dim1, y = 0, colour = party)) +
    geom_jitter(height = 0.3, alpha = 0.6, size = 1.8) +
    geom_text_repel(
      data = scores_df |> slice_min(dim1, n = 4) |>
                          bind_rows(slice_max(scores_df, dim1, n = 4)),
      aes(label = legislator_name), size = 2.8, max.overlaps = 15
    ) +
    scale_colour_manual(values = party_colours, na.value = "grey60") +
    labs(title = title, x = "Dimension 1", y = NULL, colour = "Party") +
    theme(axis.text.y  = element_blank(),
          axis.ticks.y = element_blank(),
          legend.position = "bottom")
}

p_raw_strip <- make_strip_plot(svd_raw_scores, "Raw SVD")
p_dc_strip  <- make_strip_plot(svd_dc_scores,  "Double-centred SVD")

p_raw_strip + p_dc_strip + plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

**Comparison:** [*Which representation has better face validity? Does the
double-centred version reveal within-party heterogeneity that the raw SVD
masks? Justify which you prefer and why.*]

---

# Task 3: 2PL IRT Model in brms

## Prepare long-format data for Stan

The 2PL model is fit on the **observed** (non-missing) votes only.
Legislator and bill IDs must be integer indices for Stan.

```{r irt-data}
votes_irt <- df |>
  filter(!is.na(vote)) |>
  mutate(
    legislator = as.integer(as.factor(legislator_id)),
    bill       = as.integer(as.factor(bill_id))
  ) |>
  select(vote, legislator, bill, legislator_id, bill_id, party)

cat("Observations:", nrow(votes_irt), "\n")
cat("Legislators: ", n_distinct(votes_irt$legislator), "\n")
cat("Bills:       ", n_distinct(votes_irt$bill), "\n")
```

## Specify the 2PL formula (Bürkner 2021)

The 2PL model is:
$$\Pr(X_{ij} = 1 \mid \theta_i, \alpha_j, \beta_j) = \sigma(\alpha_j + \beta_j \theta_i)$$

where $\theta_i$ is the legislator's ideal point, $\alpha_j$ is the bill
intercept (related to difficulty), and $\beta_j$ is the bill slope
(discrimination).  All three are treated as random effects in brms.

```{r irt-formula}
formula_2pl <- bf(
  vote ~ alpha + beta * theta,
  alpha ~ 1 + (1 | bill),        # per-bill intercepts (difficulty)
  beta  ~ 1 + (1 | bill),        # per-bill slopes    (discrimination)
  theta ~ 1 + (1 | legislator),  # per-legislator ideal points
  nl    = TRUE
)
```

## Weakly informative priors

```{r irt-priors}
prior_2pl <-
  prior(normal(0, 5), nlpar = "alpha") +           # bill intercepts
  prior(normal(0, 1), nlpar = "beta",  lb = 0) +   # discrimination ≥ 0
  prior(normal(0, 1), nlpar = "theta")              # ideal points ~ N(0,1)
```

> **Note on the sign constraint:** The lecture notes that political scientists
> sometimes allow $\beta_j < 0$ (bills phrased in the opposite direction).
> The `lb = 0` prior can be removed if you prefer unsigned discrimination.
> Remove `lb = 0` and let the sampler explore both sides if you go that route.

## Fit the model

Two options are provided.  Try MCMC first; fall back to variational inference
if sampling is too slow on your hardware.

```{r fit-2pl}
#| eval: false   # set to true once you have the data

# ── Option A: MCMC (1 chain, 1000 iterations — lecture-approved minimum) ──
fit_2pl <- brm(
  formula  = formula_2pl,
  data     = votes_irt,
  family   = bernoulli(link = "logit"),
  prior    = prior_2pl,
  chains   = 1,
  iter     = 1000,
  warmup   = 500,
  seed     = 42,
  file     = "models/fit_2pl_mcmc",
  backend  = "rstan"
)

# ── Option B: Variational inference (much faster, note caveat) ──────────────
# fit_2pl <- brm(
#   formula   = formula_2pl,
#   data      = votes_irt,
#   family    = bernoulli(link = "logit"),
#   prior     = prior_2pl,
#   algorithm = "meanfield",   # or "fullrank" for a richer approximation
#   seed      = 42,
#   file      = "models/fit_2pl_vi",
#   backend   = "rstan"
# )
# NOTE: VI used due to compute constraints — results are approximate.
```

## Extract ideal points and plot

```{r plot-irt}
#| eval: false   # set to true once the model is fitted

# Extract posterior means of theta (ideal points)
theta_draws <- as_draws_df(fit_2pl) |>
  select(starts_with("r_legislator__theta")) |>
  summarise(across(everything(), mean)) |>
  pivot_longer(everything(),
               names_to  = "param",
               values_to = "theta_mean") |>
  mutate(
    legislator = str_extract(param, "\\d+") |> as.integer()
  )

ideal_points <- votes_irt |>
  distinct(legislator, legislator_id, party) |>
  left_join(theta_draws, by = "legislator") |>
  left_join(leg_meta, by = "legislator_id")

# Strip plot of ideal points
ggplot(ideal_points, aes(x = theta_mean, y = 0, colour = party)) +
  geom_jitter(height = 0.3, alpha = 0.6, size = 1.8) +
  geom_text_repel(
    data = ideal_points |> slice_min(theta_mean, n = 5) |>
                           bind_rows(slice_max(ideal_points, theta_mean, n = 5)),
    aes(label = legislator_name), size = 3, max.overlaps = 20
  ) +
  scale_colour_manual(values = party_colours, na.value = "grey60") +
  labs(
    title  = "2PL IRT — Legislator Ideal Points (θ)",
    x      = "Posterior mean of θ (ideal point)",
    y      = NULL,
    colour = "Party"
  ) +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())
```

**Dimension interpretation:** [*Once the plot is produced, describe what the
first dimension captures.  Is it a left–right ideological axis?  A
government vs. opposition divide?  An economic vs. social dimension?
Name two or three legislators at each extreme and explain why they sit there.*]

---

# Comparison and Conclusions

```{r comparison}
#| eval: false   # set to true after all three models are fitted

# Side-by-side scatter: SVD dim1 vs. IRT theta
# (after aligning signs — SVD and IRT sign conventions may differ)

svd_vs_irt <- svd_raw_scores |>
  select(legislator_id, svd_raw = dim1) |>
  left_join(
    svd_dc_scores |> select(legislator_id, svd_dc = dim1),
    by = "legislator_id"
  ) |>
  left_join(
    ideal_points |> select(legislator_id, irt = theta_mean),
    by = "legislator_id"
  )

p_compare1 <- ggplot(svd_vs_irt, aes(x = svd_raw, y = irt)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE, colour = "steelblue") +
  labs(title = "Raw SVD vs. 2PL IRT", x = "SVD dim 1", y = "IRT θ")

p_compare2 <- ggplot(svd_vs_irt, aes(x = svd_dc, y = irt)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE, colour = "tomato") +
  labs(title = "Double-centred SVD vs. 2PL IRT", x = "SVD dim 1 (d-c)", y = "IRT θ")

p_compare1 + p_compare2
```

[*Write 2–4 sentences summarising: (1) whether all three methods recover a
similar ordering of legislators, (2) what the first dimension represents, and
(3) which method you find most convincing and why.*]

---

# Session Info

```{r session-info}
sessionInfo()
```
